Project Diary

### 2014-06-11 thesis/price_models.tex
Move matlab files to a separate folder.

\chapter{Return Models}\label{chp:PriceModels}
In this chapter we review some of the discrete-time return models and
fit them to intraday returns. The intention is to compare these models
in terms of forecast accuracy and to understand their statistical
properties. In the following we first describe these models briefly,
then in

% \S\ref{sec:Garch_model} reviews the GARCH model using the Nordea
% 15-minute returns as an example. \S\ref{sec:SLV_model} examines the
% stochastic log-volatility model and discusses the possibility of
% describing the residuals of the log-volotility series using the
% Johnson Su distribution. The same Nordea return series is fitted as an
% example. \S\ref{sec:XieCalc} derives the unconditional distribution
% function of the stochastic log-volotility model and compares the
% analytic results to the 30-minute returns of a number of Swedish
% stocks.

@@

The justification of modeling return series as independent,
identically distributed Gaussian variates comes from imagining the
price process $S(t)$ as a geometric Brownian motion whose increment
$\sigma dw_t$ at each time step is independent and scales as $\sigma
\sqrt{dt}$. Then, by adding a drift term $\mu dt$ that represents some
deterministic trend in the price process, one can express the
price $S(t)$ as a stochastic differential equation:
\begin{eqnarray*}
  dS &=& S\mu dt + S\sigma dw_t \\
  d(\ln S) &=& (\mu - \frac{1}{2} \sigma^2)dt + \sigma dw_t \\
\end{eqnarray*}
It follows from this equation that $\ln S(t) - \ln S_0$, where $S_0$
is the price at time 0, has Gaussian distribution with mean $(\mu -
\sigma^2/2)t$ and variance $\sigma^2 t$. Therefore, in a discrete-time
model, where the length of each time step $\Delta t$ is fixed, the
return $r_t = \ln S(t + \Delta t) - \ln S(t)$ is assumed to be
Gaussian distributed and have mean and variance that are functions of
$\Delta t$.

The picture depicted above is of course overly simplified, and the
distribution of returns is not really Gaussian. Nevertheless, the
assumption of Gaussian distributed returns underlies such important
theories as Black and Scholes theory of option pricing. Thus, albeit
inaccurate in describing very large returns, the Gaussian distribution
as a returns model shall not be forgotten. In the next few sections,
we discuss some more realistic returns models.

\item{\bf GARCH models}

``GARCH'' is the acronym for ``Generalized Autoregressive Conditional
Heteroscedasticity''. A GARCH(p, q) model is defined by the following
equation system \cite{Bollerslev86}:
\begin{equation}
  \label{eq:garch_def}
  \begin{aligned}
    r_t &= \mu_t + \epsilon_t \\
  \end{aligned}
\end{equation}
where $p$ and $q$ are constant integers; $\mu(t)$ is the mean process
and considered a small constant for intraday returns. $\epsilon_t$ is
called the innovation of the return and is a variate whose conditional
distribution \footnotemark is Gaussian and has variance $\sigma^2_t$:
\begin{eqnarray*}
  \sigma_t^2 &=& \alpha_0 + \sum_{i=1}^q \alpha_i \epsilon_{t-i}^2 +
  \sum_{i=1}^p \beta_i \sigma_{t-i}^2
\end{eqnarray*}
\footnotetext{
By conditional distribution we mean the distribution conditional on
the variate's history: The conditional distribution at time $t$ means
the distribution conditional on the history up to time $t-1$
in the discrete-time setting.
}

It can be shown that the autocorrelation function of $\epsilon_t^2$ is
given by \cite{Bollerslev87}:
\begin{eqnarray*}
\text{corr}(\epsilon_t^2, \epsilon^2_{t-n}) &=& \sum_{i=1} ^{p \vee q}
(\alpha_i + \beta_i) \rho_{n-i} \text{ for $n > p$}
\end{eqnarray*}
where $\alpha_i$ with $i > q$ and $\beta_i$ with $i > p$ are taken as
zeros. $p \vee q$ denotes the maximum of p and q. From these
equations, it is clear that the partial autocorrelation function cuts
off at $\max(p, q)$.

@@

For the purpose of intraday returns, we specify the stochastic
volatility model as
\begin{eqnarray}
   r_{t, t-h} &=& \ln S_{t} - \ln S_{t-h} \nonumber \\
   r_{t, t-h} &=& \mu + \sigma_{t, t-h} b_t \label{eq:SLV_spec}
\end{eqnarray}
where $S_t$ is the price of the asset at time $t$; $b_t \sim N(0,
1)$; $r_{t, t-h}$ is the return over the time interval $[t-h, t]$.
For simplicity, in the rest of this chapter we shall just write
$t$ for the subscript ``t, t-h'', since the time interval $h$ is fixed
for each time series and is a known constant.

Andersen and Bollerslev et al proved the following in
\cite{Andersen03} (theorem 2):
\begin{equation}
  \label{eq:normal_r}
  r_t|\mathcal{F}_{t-h} \sim N(\int_{0}^h \mu_{t-h+s} ds, \int_{0}^h
  \sigma_{t-h+s}^2 ds)
\end{equation}
In plain words, given all the information up to time $t-h$, the
distribution of $r_t$ is Gaussian with integrated mean and variance.
The variance of this conditional distribution, $\int_{0}^h
\sigma_{t-h+s}^2 ds$, can be approximated by \cite{Protter05,
  Andersen03}:
\begin{equation}
  \label{eq:rv_def}
  \int_{0}^h \sigma_{t-h+s}^2 ds = \sum_{k=1}^n \left(
    S_{t-h+kh/n} - S_{t-h+(k-1)h/n} \right)^2
\end{equation}
where $n$ is a chosen constant. The square root of the right hand side
of equation \ref{eq:rv_def} is the realized volatility, call it
$\hat{\sigma}_t$.

@@

\end{enumerate}

To compare GARCH and stochastic log-volatility models at the face of
intraday returns, we study the {\it Nordea Bank} 15-minute returns
during the period 2012/01/16 - 2012/04/20 in \S\ref{chp:nordea_15min},
and {\it Volvo B} 30-minute returns during 2013/10/10 - 2014/04/04 in
\S\ref{sec:volvo}. Another 5 intraday series are also studied and
collected in appendix \ref{chp:appendix2}.

\section{Case Study: Nordea 15-minute Returns}
\label{chp:nordea_15min}
In this section we investigate the Nordea 15-minute returns sampled
during the period 2012/01/16 - 2012/04/20. In total, these amount to
2022 returns. We use the first 80\% (1617) for model estimation and  the
remaining 20\% (405) for comparing with model forcasts. In section
\ref{sec:nordea_15min_garch} we study the series with a GARCH model 
and in section \ref{sec:nordea_15min_arima} we study it with a
stochastic volatility model.

\subsection{GARCH Model}\label{sec:nordea_15min_garch}
First of all, to determine that the volatilities are auto-correlated
and/or that the squared returns and the volatilities are correlated, and
hence a GARCH(p, q) model is appropriate for the return series under
investigation, we plot the auto-correlation function (ACF) of the
squared returns. This is shown in figure
\ref{fig:nordea_15min_acf}. If the aforementioned features are absent
from the series, the auto-correlations are expected to be Gaussian
distributed with mean 0 and variance 1/T and hence mostly reside
within the confidence bounds set by $\pm 2/\sqrt{T}$
\cite{Bollerslev86, Bollerslev87}.

Clearly this is not the case in figure \ref{fig:nordea_15min_acf} ---
the first 5 auto-correlations are rather significant. Moreover, figure
\ref{fig:nordea_15min_vlt_acf} shows even more clearly that the
conditional variances of the series are correlated. These observations
suggest a GARCH(p, q) model can be appropriate.
\begin{figure}[htb!]
  \centering
  \subfigure[]{
    \includegraphics[scale=0.4, clip=true, trim=90 257 105
    220]{../pics/nordea_15min_acf.pdf}
    \label{fig:nordea_15min_acf}
  }
  \subfigure[]{
    \includegraphics[scale=0.4, clip=true, trim=90 259 104
    218]{../pics/nordea_15min_vlt_acf.pdf}
    \label{fig:nordea_15min_vlt_acf}
  }
  \caption{\small \it \ref{fig:nordea_15min_acf}: Auto-correlations
    (ACF) of the squared returns; \ref{fig:nordea_15min_vlt_acf}:
    Auto-correlations (ACF) of squared realized volatilities
    ($\hat{\sigma}^2_t$). The blue lines are confident bounds set at
    $\pm 2/\sqrt{T}$.}
\end{figure}

Starting with a GARCH(1,1) model and taking advantage of the knowledge
that the log-volatility $\ln \sigma_t$ has seasonality $s=33$, we fit
to the return series a GARCH(33, 33) model, limiting to lags 1 and 33
for both ARCH and GARCH parameters.
\begin{eqnarray*}
  r_t &=& \mu + \epsilon_t \\
  \epsilon_t &=& \sigma_t z_t \\
  \sigma^2_t &=& \alpha_0 + \alpha_1 \epsilon^2_{t-1} + \alpha_s
  \epsilon^2_{t-s} + \beta_1 \sigma^2_{t-1} + \beta_s \sigma^2_{t-s}
\end{eqnarray*}
Via maximum likelihood estimation, parameter values listed in table
\ref{tab:nordea_15min_garch} are obtained.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    Parameter & $\alpha_0$ & $\alpha_1$ & $\alpha_s$ & $\beta_1$ &
    $\beta_s$ \\
    \hline
    Value & $4.7833 \times 10^{-7}$ & 0.1600 & 0.0667 & 0.6846 &
    0.0342 \\
    \hline
  \end{tabular}
  \caption{\small \it GARCH model parameters}
  \label{tab:nordea_15min_garch}
\end{table}

\subsection{Stochastic Volatility Model}\label{sec:nordea_15min_arima}
For the Nordea Bank 15-minute returns under consideration, it can be
verified that the square root of the sum of squared 30-second returns
makes a good proxy for the volatility. This can be seen from the
probability plot of $z_t = (r_t - \E(r_t))/\hat{\sigma}_t$ (figure
\ref{fig:nordea_bank_15min_z_prob}), i.e. the quotient of the 
15-minute returns over the volatility proxy.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.4, clip=true, trim=80 258 104
    220]{../pics/nordea_bank_15min_z_prob.pdf}
  \caption{\small \it{Probability plot of $z_t =
      (\epsilon_t-\E(\epsilon_t))/\sigma_t$. $\epsilon_t$ are derived
      from Nordea Bank 15min returns while $\sigma_t$ are realized
      volatilities calculated
      using 30s returns within each 15min interval. Horizontal axis:
      $z_t$}; Vertical axis: cummulative probability function (CDF) of
    $z_t$, arranged on such a scale that the CDF of the standard
    Gaussian is a straight line.}
  \label{fig:nordea_bank_15min_z_prob}
\end{figure}
% In addition, one can see
% from figure \ref{fig:nordea_15min_quotient_acf} and
% \ref{fig:nordea_15min_quotient_squared_acf} that there is essentially
% no auto-correlation in the $z_t$ or the $z_t^2$ series.
% \begin{figure}[htb!]
%   \centering
%   \subfigure[ACF of $z_t$]{
%     \includegraphics[scale=0.4, clip=true, trim=95 236 118
%     200]{../pics/nordea_15min_quotient_acf.pdf}
%     \label{fig:nordea_15min_quotient_acf}
%   }
%   \subfigure[ACF of $z_t^2$]{
%     \includegraphics[scale=0.4, clip=true, trim=95 236 118
%     200]{../pics/nordea_15min_quotient_squared_acf.pdf}
%     \label{fig:nordea_15min_quotient_squared_acf}
%   }
%   \caption{\small \it Nordea 15min $z_t$ and $z_t^2$ ACF.}
% \end{figure}

Andersen and Bollerslev et al reported that, for the exchange
rates between Deutch mark, yen and dollar, $\ln \sigma_t$ is
gaussian distributed \cite{Andersen03}. This is, however, not the case
for our series in question. In fact, in our case, $\ln \sigma_t$ is
right skewed (skewness 0.3342) and leptokurtic (kurtosis 
6.1006). See figure \ref{fig:nordea_15min_logvol_prob}
\begin{figure}[htb!]
  \centering
  %\vspace{-15mm}
  \includegraphics[scale=0.4, clip=true, trim=80 223 107
  4]{../pics/nordea_15min_logvol_prob.pdf}
  \caption{\small \it Probability plot of Nordea 15min
    $\ln\sigma_t$ unconditional distribution}
  \label{fig:nordea_15min_logvol_prob}
\end{figure}
Moreover, the series of $\ln\sigma_t$ shows long-lasting and
periodic autocorrelations with an apparent period of 33 (see figure
\ref{fig:nordea_15min_logvol_acf}). This suggests the series may be
described by a seasonal ARIMA model\footnote{ARIMA ---
  Integrated Autoregressive Moving Average} model. This we first
simplify the series by differencing \cite{BoxJenkins94}:
\[
w_t = (1-B)(1-B^s)\ln\sigma_t
\]
where $B$ is the back-shift operator and $s=33$ is the seasonality.

The autocorrelation function of the differenced process $w_t$, as
shown in figure \ref{fig:nordea_15min_w_acf}, clearly points to a seasonal
moving-average model: There are only 4 non-zero autocorrelations in
the plot, located at lags 1, 32, 33, 34, respectively; furthermore,
the two at 32 and 34 are approximately equal. Thus we can write down
the model as
\begin{eqnarray}
  w_t &=& (1 - \theta B)(1 - \Theta B^s) y_t \label{eq:nordea_w}
\end{eqnarray}
where $\theta$ and $\Theta$ are parameters to be determined and $y_t$
is a noise process with constant variance $\sigma_y^2$ and mean
0. $y_t$ is often refered to as the residuals.
\begin{figure}[htb!]
  \centering
  \subfigure[ACF of $\ln\sigma_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_logvol_acf.pdf}
    \label{fig:nordea_15min_logvol_acf}
  }
  \subfigure[ACF of $w_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_w_acf.pdf}
    \label{fig:nordea_15min_w_acf}
  }
  \caption{\small \it Nordea 15min $\ln\sigma_t$ and $w_t$
    autocorrelations}
  \label{fig:nordea1_15min_acf}
\end{figure}

The above seasonal moving average model has the following
autocovariance structure \cite{BoxJenkins94}:
\begin{eqnarray*}
  \gamma_0 &=& \sigma_y^2 (1 + \theta^2)(1 + \Theta^2) \\
  \gamma_1 &=& -\sigma_y^2\theta(1 + \Theta^2) \\
  \gamma_s &=& -\sigma_y^2\Theta(1 + \theta^2) \\
  \gamma_{s+1} &=& \gamma_{s-1}\;=\;\sigma_y^2\theta\Theta
\end{eqnarray*}
These equations together with the measured autocorrelations make
possible an initial estimate of the parameters $\theta$ and $\Theta$:
\begin{eqnarray*}
  {\varrho_{s+1}/\varrho_s} &=& {\gamma_{s+1}/\gamma_s} \;=\; -{\theta \over
    1 + \theta^2} \\
  {\varrho_{s+1}/\varrho_1} &=& {\gamma_{s+1}/\gamma_1} \;=\; -{\Theta \over
    1 + \Theta^2} \\
\end{eqnarray*}
Substituting in the measured values shown in table
\ref{tab:nordea_15min_w_acf},
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    $\varrho_1$ & $\varrho_{s-1}$ & $\varrho_s$ & $\varrho_{s+1}$ \\
    \hline
    -0.4703 &  0.2053 & -0.4564 &  0.2212 \\
    \hline
  \end{tabular}
  \caption{\small \it Nordea 15min $w_t$ autocorrelations}
  \label{tab:nordea_15min_w_acf}
\end{table}
we get
\begin{eqnarray*}
  \theta &=& 0.6890 \\
  \Theta &=& 0.6378
\end{eqnarray*}
Among the two roots of each of the 2nd order equations in the above,
we have chosen the one in the range $(-1, 1)$ so as to ensure
invertibility of the model \cite{BoxJenkins94}.

>>>>

\chapter{Return Models}\label{chp:PriceModels}
In this chapter we review some of the discrete-time return models and
fit them to intraday returns. The intention is to compare these models
in terms of forecast accuracy and to understand their statistical
properties. In the following we first describe these models briefly,
then in section \ref{chp:nordea_15min}, section \ref{sec:volvo_30} and appendix
\ref{chp:appendix2} we describe how the \gls{garch} and the \gls{sv} models
are fitted to intraday returns and compare their forecasts. In section
\ref{sec:XieCalc} we calculate the unconditional distribution
functions of the \gls{sv} model.

% \gls{frt} is good.
% section \ref{sec:Garch_model} reviews the \gls{garch} model using the Nordea
% 15-minute returns as an example. Section \ref{sec:SLV_model} examines the
% stochastic log-volatility model and discusses the possibility of
% describing the residuals of the log-volotility series using the
% Johnson Su distribution. The same Nordea return series is fitted as an
% example. Section \ref{sec:XieCalc} derives the unconditional distribution
% function of the stochastic log-volotility model and compares the
% analytic results to the 30-minute returns of a number of Swedish
% stocks.

@@

The justification of modeling return series as independent,
identically distributed Gaussian variates comes from imagining the
price process $S(t)$ as a geometric Brownian motion whose increment
$\sigma dw_t$ at each time step is independent and scales as $\sigma
\sqrt{dt}$. Then, by adding a drift term $\mu dt$ that represents some
deterministic trend in the price process, one can express the
price $S(t)$ as a stochastic differential equation:
\begin{eqnarray*}
  dS &=& S\mu dt + S\sigma dw_t \\
\end{eqnarray*}
Then by It\^o's lemma, a stochastic differential equation for the
logarithmic price $\ln S$ can be obtained
\begin{eqnarray*}
  d(\ln S) &=& (\mu - \frac{1}{2} \sigma^2)dt + \sigma dw_t \\
\end{eqnarray*}
It follows from this equation that $\ln S(t) - \ln S_0$, where $S_0$
is the price at time 0, has Gaussian distribution with mean $(\mu -
\sigma^2/2)t$ and variance $\sigma^2 t$ \cite{Bernt2000}. Therefore,
in a discrete-time model, where the length of each time step $\Delta
t$ is fixed, the return $r_t = \ln S(t) - \ln S(t - \Delta t)$ is
assumed to be Gaussian distributed and have mean and variance that are
functions of $\Delta t$.

The picture depicted above is of course overly simplified, and the
distribution of returns is not really Gaussian. Nevertheless, the
assumption of Gaussian distributed returns underlies such important
theories as Black and Scholes theory of option pricing. Thus, albeit
inaccurate in describing very large returns, the Gaussian distribution
as a return model shall not be forgotten. In the next few sections,
we discuss some more realistic return models.

\item{\bf \gls{garch} models}

``GARCH'' is the acronym for ``Generalized Autoregressive Conditional
Heteroscedasticity''. A \gls{garch}(p, q) model is defined by the following
equation system \cite{Bollerslev86}:
\begin{equation}
  \label{eq:garch_def}
  \begin{aligned}
    r_t &= \mu_t + \epsilon_t \\
  \end{aligned}
\end{equation}
where $r_t$ is the return as mentioned earlier; $p$ and $q$ are
constant integers; $\mu_t$ is the mean process and considered a small
constant for intraday returns. $\epsilon_t$ is called the innovation
of the return and is a variate whose conditional distribution
\footnotemark is Gaussian and has variance $\sigma^2_t$:
\begin{eqnarray*}
  \sigma_t^2 &=& \alpha_0 + \sum_{i=1}^q \alpha_i \epsilon_{t-i}^2 +
  \sum_{i=1}^p \beta_i \sigma_{t-i}^2
\end{eqnarray*}
\footnotetext{
By conditional distribution we mean the distribution conditional on
the variate's history: The conditional distribution at time $t$ means
the distribution conditional on the history up to time $t-1$
in the discrete-time setting.
}

It can be shown that the autocorrelation function $\varrho_n =
\text{corr}(\epsilon_t, \epsilon_{t-n})$ for $n > \text{max}(p,q)$ of
$\epsilon_t^2$ is given by \cite{Bollerslev87}:
\begin{eqnarray*}
\varrho_n &=& \sum_{i=1} ^{p \vee q}
(\alpha_i + \beta_i) \varrho_{n-i} \text{ for $n > p$}
\end{eqnarray*}
where $\alpha_i$ with $i > q$ and $\beta_i$ with $i > p$ are taken as
zeros. $p \vee q$ denotes the maximum of p and q. From these
equations, it is clear that the partial autocorrelation function cuts
off at $\max(p, q)$.

Here we note that, in the \gls{garch} model, the volatility $\sigma_t$ is
$\mathcal{F}_{t-1}$ measurable: given the history up to time $t-1$,
$\sigma_t$ is a deterministic quantity. In contrast, \gls{sv} models,
which we discuss shortly, treat $\sigma_t$ as a random variable even
given the same history.

@@

For the purpose of intraday returns, we specify the \gls{sv} model as
\begin{eqnarray}
   r_{t, t-h} &=& \ln S_{t} - \ln S_{t-h} \nonumber \\
   r_{t, t-h} &=& \mu + \sigma_{t, t-h} b_t \label{eq:SLV_spec}
\end{eqnarray}
where $S_t$ is the price of the asset at time $t$; $b_t \sim N(0,
1)$; $r_{t, t-h}$ is the return over the time interval $[t-h, t]$.
For simplicity, in the rest of this chapter we shall just write
$t$ for the subscript ``t, t-h'', since the time interval $h$ is fixed
for each time series and is a known constant.

Andersen et al proved the following in \cite{Andersen03} (theorem 2):
\begin{equation}
  \label{eq:normal_r}
  r_t|\mathcal{F}_{t-h} \sim N(\int_{0}^h \mu_{t-h+s} ds, \int_{0}^h
  \sigma_{t-h+s}^2 ds)
\end{equation}
In plain words, conditional on the information up to time $t-h$, the
distribution of $r_t$ is Gaussian with integrated mean and variance.
The variance of this conditional distribution, $\int_{0}^h
\sigma_{t-h+s}^2 ds$, can be approximated by \cite{Protter05,
  Andersen03}:
\begin{equation}
  \label{eq:rv_def}
  \int_{0}^h \sigma_{t-h+s}^2 ds = \sum_{k=1}^n \left(
    \ln S_{t-h+kh/n} - \ln S_{t-h+(k-1)h/n} \right)^2
\end{equation}
where $n$ is a chosen constant. The square root of the right hand side
of equation \ref{eq:rv_def} is the realized volatility, call it
$\hat{\sigma}_t$.

@@

Once a good approximation of $\sigma_t$ has been found, it is
convenient to model $\ln\sigma_t$ so that the positivity of $\sigma_t$
is implied by construction \cite{Mikosch2009}. As is seen in later
sections, \gls{arima} models often serve well for this purpose. An
\gls{arima}(p, d, q) model, where $p,d,q$ are integers, is defined as
\cite{BoxJenkins94}
\begin{eqnarray*}
  (1 - \sum_{i=1}^p \phi_i B^i) (1 - B)^d \ln \sigma_t &=& (1 -
  \sum_{i=1}^q \theta_i B^i) y_t
\end{eqnarray*}
where $B$ is the back-shift operator such that $B x_t = x_{t-1}$ for
any time series $x_t$. $y_t$ are termed the residuals of the
model, $d$ is the order of integration, and $p, q$  are orders of
autoregression and moving average, respectively. $\phi_i$ and
$\theta_i$ are constant parameters.

Quite often, the auto-correlation function of the time series in
question manifests periodic patterns, thus seasonal components are
added to the above model to account for the seasonality:
\begin{eqnarray*}
&& (1 - \sum_{i=1}^P \Phi_i B^{is}) (1 - \sum_{i=1}^p \phi_i B^i) (1 - B^s)^D (1 -
B)^d\ln \sigma_t \\
&& = (1 - \sum_{i=1}^Q \Theta_i B^{is}) (1 -
\sum_{i=1}^q \theta_i B^i)y_t
\end{eqnarray*}
where $D$, analogous to $d$ in the non-seasonal model, is the order of
seasonal integration; $P, Q$ are orders of seasonal autoregression and
moving average, respectively. $\Phi_i$ and $\Theta_i$ are constant
parameters. In the most general situations, the seasonal and the
non-seasonal components do not necessarily combine in the above
multiplicative fashion, so the following model is also of interest:
\begin{eqnarray*}
(1 - \sum_{i=1}^{p+P} \phi_i B^i) (1 - B^s)^D(1 - B)^d \ln
\sigma_t &=& (1 - \sum_{i=1}^{q+Q} \theta_i B^i) y_t
\end{eqnarray*}
In all the above cases, if $d = 0$ and $D = 0$, the model does not
involve integration and hence is called an \gls{arma}(p,q) model. An
even simpler case is where $q = 0$ and $Q = 0$ in addition to $d = 0$
and $D = 0$. These conditions make the model a pure \gls{ar}
process, denoted AR(p).
\end{enumerate}

To compare \gls{garch} and \gls{sv} models at the face of intraday returns,
we study the {\it Nordea Bank} 15-minute returns during the period
2012/01/16 - 2012/04/20 in section \ref{chp:nordea_15min}, 
and {\it Volvo B} 30-minute returns during 2013/10/10 - 2014/04/04 in
section \ref{sec:volvo_30}. Another 4 intraday series are also studied and
collected in appendix \ref{chp:appendix2}. We have chosen these
particular stocks because they have the largest trading volumes in the
Swedish market and hence provide the largest amount of data for
analysis. The time intervals of 15 and 30 minutes are chosen because
they are relatively short and hence give a large amount of data and
yet they are not too short for noise of market
micro-structure\footnotemark to become a concern.
\footnotetext{
Noise of market micro-structure is due to, most importantly,
the difference between the bid and asked prices (bid-ask spread), and
the discreteness of price changes.
}

\section{Case Study: Nordea 15-minute Returns}
\label{chp:nordea_15min}
In this section we investigate the Nordea 15-minute returns sampled
during the period 2012/01/16 - 2012/04/20. In total, these amount to
2022 returns. We use the first 80\% (1617) for model estimation and  the
remaining 20\% (405) for comparing with model forcasts. In section
\ref{sec:nordea_15min_garch} we study the series with a \gls{garch} model 
and in section \ref{sec:nordea_15min_arima} we study it with a
\gls{sv} model.

\subsection{GARCH Model}\label{sec:nordea_15min_garch}
When volatilities are auto-correlated or squared returns and
volatilities are correlated, a \gls{garch}(p, q) model may be
appropriate for the return series under investigation. To find out
whehter this is true in our case, we plot the \gls{acf} of the squared
returns. This is shown in figure \ref{fig:nordea_15min_acf}. If the
aforementioned features are absent from the series, the
auto-correlations are expected to be Gaussian distributed with mean 0
and variance 1/T, and hence mostly reside within the confidence bounds
set by $\pm 2/\sqrt{T}$ \cite{Bollerslev86, Bollerslev87}.

Clearly this is not the case in figure \ref{fig:nordea_15min_acf} ---
the first 5 auto-correlations are rather significant. Moreover, figure
\ref{fig:nordea_15min_vlt_acf} shows even more clearly that the
conditional variances of the series are correlated. These observations
suggest a \gls{garch}(p, q) model can be appropriate.
\begin{figure}[htb!]
  \centering
  \subfigure[]{
    \includegraphics[scale=0.4, clip=true, trim=90 257 105
    220]{../pics/nordea_15min_acf.pdf}
    \label{fig:nordea_15min_acf}
  }
  \subfigure[]{
    \includegraphics[scale=0.4, clip=true, trim=90 259 104
    218]{../pics/nordea_15min_vlt_acf.pdf}
    \label{fig:nordea_15min_vlt_acf}
  }
  \caption{\small \it \ref{fig:nordea_15min_acf}: Auto-correlations
    (ACF) of the squared returns; \ref{fig:nordea_15min_vlt_acf}:
    Auto-correlations (ACF) of squared realized volatilities
    ($\hat{\sigma}^2_t$). The blue lines are confident bounds set at
    $\pm 2/\sqrt{T}$. $T$ is the length of the time series.}
\end{figure}

Starting with a \gls{garch}(1,1) model and taking advantage of the knowledge
that the log-volatility $\ln \sigma_t$ has seasonality $s=33$ (see
figure \ref{fig:nordea_15min_logvol_acf}), we fit to the return series
a \gls{garch}(33, 33) model, limiting to lags 1 and 33 for both ARCH and
\gls{garch} parameters.
\begin{eqnarray*}
  r_t &=& \mu + \epsilon_t \\
  \epsilon_t &=& \sigma_t z_t \\
  \sigma^2_t &=& \alpha_0 + \alpha_1 \epsilon^2_{t-1} + \alpha_s
  \epsilon^2_{t-s} + \beta_1 \sigma^2_{t-1} + \beta_s \sigma^2_{t-s}
\end{eqnarray*}
where the mean process of $r_t$, denoted $\mu_t$ earlier, is simplified
to a mere constant $\mu$ owing to its smallness. by \gls{mle},
parameter values listed in table \ref{tab:nordea_15min_garch} are obtained.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    Parameter & $\alpha_0$ & $\alpha_1$ & $\alpha_s$ & $\beta_1$ &
    $\beta_s$ \\
    \hline
    Value & $4.7833 \times 10^{-7}$ & 0.1600 & 0.0667 & 0.6846 &
    0.0342 \\
    \hline
  \end{tabular}
  \caption{\small \it GARCH model parameters}
  \label{tab:nordea_15min_garch}
\end{table}

\subsection{Stochastic Volatility Model}\label{sec:nordea_15min_arima}
For the Nordea Bank 15-minute returns under consideration, it can be
verified that the square root of the sum of squared 30-second returns
makes a good proxy for the volatility. This can be seen from the
probability plot of $z_t = (r_t - \E(r_t))/\hat{\sigma}_t$ (figure
\ref{fig:nordea_bank_15min_z_prob}), i.e. the quotient of the 
15-minute returns over the volatility proxy.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.4, clip=true, trim=80 258 104
    220]{../pics/nordea_bank_15min_z_prob.pdf}
  \caption{\small \it Probability plot of $z_t =
    (r_t-\E(r_t))/\sigma_t$. $\epsilon_t$ are derived 
      from Nordea Bank 15min returns while $\sigma_t$ are realized
      volatilities calculated using 30s returns within each 15min
      interval. Horizontal axis: $z_t$; Vertical axis: CDF of
    $z_t$, arranged on such a scale that the CDF of the standard
    Gaussian is a straight line.}
  \label{fig:nordea_bank_15min_z_prob}
\end{figure}
% In addition, one can see
% from figure \ref{fig:nordea_15min_quotient_acf} and
% \ref{fig:nordea_15min_quotient_squared_acf} that there is essentially
% no auto-correlation in the $z_t$ or the $z_t^2$ series.
% \begin{figure}[htb!]
%   \centering
%   \subfigure[ACF of $z_t$]{
%     \includegraphics[scale=0.4, clip=true, trim=95 236 118
%     200]{../pics/nordea_15min_quotient_acf.pdf}
%     \label{fig:nordea_15min_quotient_acf}
%   }
%   \subfigure[ACF of $z_t^2$]{
%     \includegraphics[scale=0.4, clip=true, trim=95 236 118
%     200]{../pics/nordea_15min_quotient_squared_acf.pdf}
%     \label{fig:nordea_15min_quotient_squared_acf}
%   }
%   \caption{\small \it Nordea 15min $z_t$ and $z_t^2$ ACF.}
% \end{figure}

Andersen and Bollerslev et al reported that, for the exchange
rates between Deutch mark, yen and dollar, $\ln \sigma_t$ is
Gaussian distributed \cite{Andersen03}. This is, however, not the case
for our modestly sized series. In fact, in our case, $\ln \sigma_t$ is
right skewed (skewness 0.3342) and leptokurtic (kurtosis 6.1006). See
figure \ref{fig:nordea_15min_logvol_prob}
\begin{figure}[htb!]
  \centering
  %\vspace{-15mm}
  \includegraphics[scale=0.4, clip=true, trim=80 223 107
  4]{../pics/nordea_15min_logvol_prob.pdf}
  \caption{\small \it Probability plot of Nordea 15min log-volatility
    $\ln\sigma_t$ unconditional distribution}
  \label{fig:nordea_15min_logvol_prob}
\end{figure}
Moreover, the series of $\ln\sigma_t$ shows long-lasting and
periodic autocorrelations with an apparent period of 33 (see figure
\ref{fig:nordea_15min_logvol_acf}). This suggests the series may be
described by a seasonal \gls{arima} model. Thus we first simplify the
series by differencing \cite{BoxJenkins94}:
\begin{equation}
  \label{eq:differenced_lv}
  w_t = (1-B)(1-B^s)\ln\sigma_t  
\end{equation}
where $B$ is the back-shift operator\footnote{For example, $B\,x_t =
  x_{t-1}$} and $s=33$ is the seasonality.

The autocorrelation function of the differenced process $w_t$, as
shown in figure \ref{fig:nordea_15min_w_acf}, clearly points to a seasonal
moving-average model: There are only 4 non-zero autocorrelations in
the plot, located at lags 1, 32, 33, 34, respectively; furthermore,
the two at 32 and 34 are approximately equal. Thus we can write down
the model as
\begin{eqnarray}
  w_t &=& (1 - \theta B)(1 - \Theta B^s) y_t \label{eq:nordea_w}
\end{eqnarray}
where $\theta$ and $\Theta$ are parameters to be determined and $y_t$
is a noise process with constant variance $\sigma_y^2$ and mean
0. $y_t$ is often refered to as the residuals.
\begin{figure}[htb!]
  \centering
  \subfigure[ACF of log-volatility ($\ln\sigma_t$)]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_logvol_acf.pdf}
    \label{fig:nordea_15min_logvol_acf}
  }
  \subfigure[ACF of differenced log-volatility ($w_t$)]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_w_acf.pdf}
    \label{fig:nordea_15min_w_acf}
  }
  \caption{\small \it Auto-correlations of Nordea 15min log-volatility
    ($\ln\sigma_t$) and differenced log-volatility ($w_t$).}
  \label{fig:nordea1_15min_acf}
\end{figure}
